---
title: Pairwise Graph Alignment with cone_align
output: rmarkdown::html_vignette
vignette: '%\VignetteIndexEntry{Pairwise Graph Alignment with cone_align} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}'
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`cone_align()` aligns two graph domains by alternating between spectral
embeddings, orthogonal Procrustes rotations, and linear assignment.  This
vignette demonstrates the workflow on the shared
`alignment_benchmark` dataset so that results are directly comparable to other
alignment methods in the package.

## Setup

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
library(manifoldalign)
library(multidesign)
library(multivarious)
library(tibble)
library(dplyr)
library(ggplot2)
```

## Benchmark Pair

We use the first two domains from `alignment_benchmark`.  Each domain contains
80 nodes, a 4-dimensional feature matrix, and a `design` data frame with the
latent class label (`condition`).

```{r load-data}
alignment_benchmark <- manifoldalign::alignment_benchmark
pair_domains <- alignment_benchmark$domains[1:2]

pair_multidesign <- lapply(pair_domains, function(dom) {
  multidesign::multidesign(dom$x, dom$design)
})

pair_names <- names(pair_multidesign)
node_counts <- vapply(pair_multidesign, function(dom) nrow(dom$x), integer(1))

hd_pair <- multidesign::hyperdesign(pair_multidesign)
```

## Running `cone_align`

```{r run-cone-align}
cone_fit <- cone_align(
  hd_pair,
  ncomp = 8,
  sigma = 0.9,
  lambda = 0.05,
  solver = "linear",
  max_iter = 40,
  tol = 1e-3
)

str(cone_fit, max.level = 1)
```

### Assignment Quality

`cone_align()` returns a permutation vector (`assignment`) that maps nodes in the
second domain to nodes in the first (reference) domain.  The latent class labels
were not used during optimisation—they are only consulted here as a diagnostic to
see how often the unsupervised alignment preserves the original classes.

```{r assignment-quality}
assignment <- cone_fit$assignment
ref_labels <- pair_multidesign[[1]]$design$condition
cmp_labels <- pair_multidesign[[2]]$design$condition

condition_accuracy <- mean(ref_labels[assignment] == cmp_labels)
identity_fraction <- mean(assignment == seq_along(assignment))

tibble(
  metric = c("Class agreement", "Exact identity"),
  value = c(condition_accuracy, identity_fraction)
)
```

### Visualising the Aligned Embeddings

The aligned embeddings (`cone_fit$s`) stack the two domains.  We visualise the
first two components and draw a handful of matching pairs to illustrate the
assignments.

```{r plot-embeddings, fig.width = 6, fig.height = 4}
# Note: class labels are only used for visual diagnostics below; `cone_align`
# operates in a completely unsupervised manner and does not see `condition`.

score_tbl <- as_tibble(as.matrix(cone_fit$s), .name_repair = "minimal")
colnames(score_tbl) <- paste0("comp", seq_len(ncol(score_tbl)))

scores <- score_tbl %>%
  mutate(
    domain = rep(pair_names, times = node_counts),
    condition = rep(ref_labels, length(pair_names))
  )

# pick 15 nodes to visualise correspondences
set.seed(123)
subset_idx <- sample(seq_len(node_counts[1]), size = 15)
segment_df <- tibble(
  x = scores$comp1[subset_idx],
  y = scores$comp2[subset_idx],
  xend = scores$comp1[node_counts[1] + assignment[subset_idx]],
  yend = scores$comp2[node_counts[1] + assignment[subset_idx]]
)

 ggplot(scores, aes(x = comp1, y = comp2, colour = condition, shape = domain)) +
  geom_point(size = 2.1, alpha = 0.85) +
  geom_segment(
    data = segment_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,
    colour = "grey60",
    linewidth = 0.4,
    alpha = 0.7
  ) +
  labs(title = "cone_align embeddings (first two components)",
       x = "Component 1", y = "Component 2") +
  theme_minimal()
```

## Summary

- `cone_align()` produces orthogonally aligned embeddings together with a node
  correspondence.
- On this benchmark pair the assignments correctly match the underlying class
  in ~`r sprintf("%.1f%%", condition_accuracy * 100)` of cases.
- The same dataset is used in the `kema` and `gpca_align` vignettes, enabling
  easy comparison across alignment techniques, while remembering that
  `cone_align()` itself remains class-agnostic.
