---
title: Multi-Domain Graph Alignment with cone_align_multiple
output: rmarkdown::html_vignette
vignette: '%\VignetteIndexEntry{Multi-Domain Graph Alignment with cone_align_multiple}
  %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}'
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`cone_align_multiple()` extends CONE-Align to three or more graph domains by
iteratively aligning each domain to a chosen reference graph.  The algorithm
repeatedly solves pairwise Procrustes/assignment problems until node
correspondences stabilise.

This vignette demonstrates the workflow on the shared
`alignment_benchmark` dataset.

## Setup

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
library(manifoldalign)
library(multidesign)
library(multivarious)
library(tibble)
library(dplyr)
library(ggplot2)
```

## Loading the Benchmark Domains

```{r load-data}
alignment_benchmark <- manifoldalign::alignment_benchmark
multi_domains <- lapply(alignment_benchmark$domains, function(dom) {
  multidesign::multidesign(dom$x, dom$design)
})

multi_names <- names(multi_domains)
node_counts <- vapply(multi_domains, function(dom) nrow(dom$x), integer(1))

hd_multi <- multidesign::hyperdesign(multi_domains)
```

## Running `cone_align_multiple`

```{r run-cone-multi}
cone_multi <- cone_align_multiple(
  hd_multi,
  ref_idx = 1,
  ncomp = 6,
  sigma = 0.9,
  lambda = 0.05,
  solver = "linear",
  max_iter = 40,
  tol = 1e-3
)

str(cone_multi, max.level = 1)
```

The reference domain (domain 1) is aligned to itself with the identity
permutation.  For the remaining domains we evaluate how often aligned pairs
share the same latent class—again, labels were not used during optimisation and
serve purely as a diagnostic.

```{r multi-accuracy}
assignments <- cone_multi$assignment
ref_labels <- multi_domains[[1]]$design$condition

accuracy_values <- vapply(seq_along(multi_names)[-1], function(idx) {
  mean(ref_labels[assignments[[idx]]] == multi_domains[[idx]]$design$condition)
}, numeric(1))

accuracy_tbl <- tibble(
  domain = multi_names[-1],
  condition_accuracy = accuracy_values
)

accuracy_tbl
```

## Visualising All Domains

```{r multi-plot, fig.width = 6, fig.height = 4}
# As with the pairwise vignette, class labels are only used for evaluation and
# plotting; `cone_align_multiple()` itself is unsupervised.

score_tbl <- as_tibble(as.matrix(cone_multi$s), .name_repair = "minimal")
colnames(score_tbl) <- paste0("comp", seq_len(ncol(score_tbl)))

scores <- score_tbl %>%
  mutate(
    domain = rep(multi_names, times = node_counts),
    condition = rep(ref_labels, length(multi_names))
  )

 ggplot(scores, aes(x = comp1, y = comp2, colour = condition, shape = domain)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = "cone_align_multiple embeddings (first two components)",
       x = "Component 1", y = "Component 2") +
  theme_minimal()
```

## Summary

- `cone_align_multiple()` aligns every domain to a shared reference using a
  sequence of pairwise CONE-Align steps.
- On this benchmark the recovered correspondences preserve class labels across
  all domains (see the accuracy table above).
- Because the same dataset powers the `kema` and `gpca_align` vignettes, you can
  now benchmark graph-vs-kernel-vs-linear alignment on identical inputs while
  recognising that the CONE variants remain unsupervised.
