---
title: Coupled Diagonalization across Modalities
output: rmarkdown::html_vignette
vignette: '%\VignetteIndexEntry{Coupled Diagonalization across Modalities} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}'
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`coupled_diagonalization()` finds a shared basis for several modalities by jointly
optimising Laplacian diagonalisation and cross-domain agreement. The routine is
unsupervised; class labels appear only in diagnostics below.

## Setup

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
library(manifoldalign)
library(multidesign)
library(multivarious)
library(tibble)
library(dplyr)
library(ggplot2)
```

## Loading the Benchmark Domains

```{r load-data}
bench <- manifoldalign::alignment_benchmark
multi_domains <- lapply(bench$domains, function(dom) {
  multidesign::multidesign(dom$x, dom$design)
})

multi_names <- names(multi_domains)
node_counts <- vapply(multi_domains, function(dom) nrow(dom$x), integer(1))

hd_multi <- multidesign::hyperdesign(multi_domains)
```

## Running Coupled Diagonalization

```{r run-cd}
cd_fit <- coupled_diagonalization(
  hd_multi,
  ncomp = 5,
  ncomp_per_domain = 12,
  mu_coupling = 1.5,
  knn = 8,
  verbose = FALSE
)

list(
  converged = cd_fit$converged,
  iterations = cd_fit$iterations,
  final_cost = cd_fit$final_cost
)
```

### Coupled Basis Diagnostics

We inspect the Frobenius norm of each coupled basis and pairwise Hilbert-Schmidt
similarities to gauge alignment strength.

```{r cd-diagnostics}
basis_norms <- purrr::imap_dfr(cd_fit$coupled_bases, function(Ai, name) {
  tibble(domain = name, frobenius_norm = sqrt(sum(Ai^2)))
})

basis_similarity <- function(A, B) sqrt(sum((t(A) %*% B)^2))

pair_indices <- combn(names(cd_fit$coupled_bases), 2, simplify = FALSE)
similarity_tbl <- purrr::map_dfr(pair_indices, function(pair) {
  tibble(
    domain_i = pair[1],
    domain_j = pair[2],
    hs_similarity = basis_similarity(cd_fit$coupled_bases[[pair[1]]],
                                     cd_fit$coupled_bases[[pair[2]]])
  )
})

basis_norms
similarity_tbl
```

### Visualising Coupled Components

```{r cd-plot, fig.width = 6, fig.height = 4}
score_tbl <- as_tibble(as.matrix(cd_fit$s), .name_repair = "minimal")
colnames(score_tbl) <- paste0("comp", seq_len(ncol(score_tbl)))

scores <- score_tbl %>%
  mutate(
    sample = seq_len(nrow(.)),
    domain = rep(multi_names, times = node_counts),
    condition = rep(bench$labels, length(multi_names))
  )

 ggplot(scores, aes(x = comp1, y = comp2, colour = condition, shape = domain)) +
  geom_point(alpha = 0.85, size = 2) +
  labs(title = "Coupled diagonalization: first two components",
       x = "Component 1", y = "Component 2") +
  theme_minimal()
```

## Summary

- `coupled_diagonalization()` produces orthogonal bases that align the domains’
  eigenvectors.
- Basis norms and Hilbert–Schmidt similarities (see above) quantify the quality of
  the joint diagonalisation.
- Because all vignettes reuse `alignment_benchmark`, you can compare coupled
  diagonalisation directly against graph-, kernel-, and OT-based approaches.
