---
title: 'Features, Correspondences, and Predictive Performance'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Features, Correspondences, and Predictive Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(params$family))
}
```

Suppose domain $d$ has an observed matrix $X_d \in \mathbb{R}^{n_d \times p_d}$.
Each **row** is a sample. Now suppose the same rows also have external feature
vectors, collected in a second matrix $F_d \in \mathbb{R}^{n_d \times q}$.
Those row-level features might come from text embeddings, anatomy, metadata, or
another modality entirely.

That distinction matters:

- columns of `X_d` are the measured variables used to fit the alignment;
- rows of `F_d` are sample-level descriptors used to define semantic
  correspondences and held-out evaluation;
- a cross-domain feature correspondence matrix between domains `a` and `b` is
  therefore `nrow(a) x nrow(b)`, not `ncol(a) x ncol(b)`.

In this vignette we will:

- build row-level feature correspondences for **three** domains;
- keep those correspondences sparse;
- fit SSMA on training rows only;
- project held-out rows with `oos_predict()`;
- score predictive performance with `cv_alignment_rows()`.

The predictive target is not "reconstruct the feature vector." It is: does a
held-out row land near rows in other domains whose **row-level features** are
similar?

## Setup

```{r libraries}
library(manifoldalign)
library(multidesign)
library(Matrix)
library(ggplot2)
```

```{r helpers, include = FALSE}
row_cosine_similarity <- function(Fi, Fj) {
  Fi <- as.matrix(Fi)
  Fj <- as.matrix(Fj)

  Fi_norm <- sqrt(rowSums(Fi * Fi))
  Fj_norm <- sqrt(rowSums(Fj * Fj))
  Fi_norm[Fi_norm <= 0] <- 1
  Fj_norm[Fj_norm <= 0] <- 1

  (Fi / Fi_norm) %*% t(Fj / Fj_norm)
}

feature_correspondence_matrix <- function(Fi,
                                          Fj,
                                          top_k = 2L,
                                          mutual = TRUE,
                                          min_similarity = -Inf) {
  sim <- row_cosine_similarity(Fi, Fj)
  n_i <- nrow(sim)
  n_j <- ncol(sim)

  top_k <- max(1L, min(as.integer(top_k), n_j))
  row_keep <- matrix(FALSE, n_i, n_j)
  for (i in seq_len(n_i)) {
    row_keep[i, utils::head(order(sim[i, ], decreasing = TRUE), top_k)] <- TRUE
  }

  keep <- row_keep
  if (isTRUE(mutual)) {
    top_k_col <- max(1L, min(as.integer(top_k), n_i))
    col_keep <- matrix(FALSE, n_i, n_j)
    for (j in seq_len(n_j)) {
      col_keep[utils::head(order(sim[, j], decreasing = TRUE), top_k_col), j] <- TRUE
    }
    keep <- keep & col_keep
  }

  if (is.finite(min_similarity)) {
    keep <- keep & (sim >= min_similarity)
  }

  Matrix::Matrix(sim * keep, sparse = TRUE)
}

feature_correspondences <- function(feature_blocks,
                                    top_k = 2L,
                                    mutual = TRUE,
                                    min_similarity = 0.5) {
  if (!is.list(feature_blocks) || length(feature_blocks) < 2L) {
    stop("`feature_blocks` must be a list with at least two matrices.", call. = FALSE)
  }

  pair_index <- utils::combn(seq_along(feature_blocks), 2L, simplify = FALSE)
  rows <- vector("list", length(pair_index))

  for (r in seq_along(pair_index)) {
    pair <- pair_index[[r]]
    Wij <- feature_correspondence_matrix(
      feature_blocks[[pair[1]]],
      feature_blocks[[pair[2]]],
      top_k = top_k,
      mutual = mutual,
      min_similarity = min_similarity
    )
    nz <- summary(Wij)

    rows[[r]] <- data.frame(
      domain_i = pair[1],
      index_i = nz$i,
      domain_j = pair[2],
      index_j = nz$j,
      weight = nz$x,
      source = "feature",
      stringsAsFactors = FALSE
    )
  }

  keep <- vapply(rows, nrow, integer(1)) > 0L
  if (!any(keep)) {
    return(data.frame(
      domain_i = integer(0),
      index_i = integer(0),
      domain_j = integer(0),
      index_j = integer(0),
      weight = numeric(0),
      source = character(0),
      stringsAsFactors = FALSE
    ))
  }

  out <- do.call(rbind, rows[keep])
  rownames(out) <- NULL
  out
}

score_row_blocks <- function(hd) {
  block_sizes <- vapply(hd, function(block) nrow(block$x), integer(1))
  end_idx <- cumsum(block_sizes)
  start_idx <- c(1L, head(end_idx, -1L) + 1L)

  out <- lapply(seq_along(block_sizes), function(i) {
    seq.int(start_idx[i], end_idx[i])
  })
  names(out) <- names(hd)
  out
}

latent_knn <- function(query, target, k = 3L) {
  query <- as.matrix(query)
  target <- as.matrix(target)

  if (requireNamespace("RANN", quietly = TRUE)) {
    return(RANN::nn2(data = target, query = query, k = k)$nn.idx)
  }

  s1 <- rowSums(query * query)
  s2 <- rowSums(target * target)
  D <- outer(s1, s2, "+") - 2 * (query %*% t(target))

  if (k == 1L) {
    return(matrix(max.col(-D, ties.method = "first"), ncol = 1L))
  }

  t(vapply(seq_len(nrow(D)), function(i) {
    utils::head(order(D[i, ], decreasing = FALSE), k)
  }, integer(k)))
}

aggregate_cv <- function(res, label) {
  data.frame(
    scenario = label,
    metric = c("top-1", "top-k"),
    value = c(
      mean(res$scores$mean_top1_similarity),
      mean(res$scores$mean_topk_similarity)
    )
  )
}
```

## Simulate three observed domains plus row-level features

We will simulate three observed domains from a shared latent state. Separately,
we will generate an external feature matrix for each domain. Those feature
vectors are attached to the **rows** of each dataset and are only used for
correspondence construction and evaluation.

```{r simulate-data}
set.seed(20260420)
n <- 72
latent_dim <- 3
feature_dim <- 5

Z <- matrix(rnorm(n * latent_dim), n, latent_dim)

A1 <- matrix(rnorm(latent_dim * 8), latent_dim, 8)
A2 <- matrix(rnorm(latent_dim * 6), latent_dim, 6)
A3 <- matrix(rnorm(latent_dim * 9), latent_dim, 9)

observed <- list(
  a = Z %*% A1 + matrix(rnorm(n * 8, sd = 0.08), n, 8),
  b = Z %*% A2 + matrix(rnorm(n * 6, sd = 0.08), n, 6),
  c = Z %*% A3 + matrix(rnorm(n * 9, sd = 0.08), n, 9)
)

feature_basis <- matrix(rnorm(latent_dim * feature_dim), latent_dim, feature_dim)
feature_signal <- Z %*% feature_basis

features_true <- list(
  a = feature_signal + matrix(rnorm(n * feature_dim, sd = 0.04), n, feature_dim),
  b = feature_signal + matrix(rnorm(n * feature_dim, sd = 0.04), n, feature_dim),
  c = feature_signal + matrix(rnorm(n * feature_dim, sd = 0.04), n, feature_dim)
)

domain_summary <- data.frame(
  domain = names(observed),
  n_rows = vapply(observed, nrow, integer(1)),
  observed_columns = vapply(observed, ncol, integer(1)),
  feature_columns = vapply(features_true, ncol, integer(1))
)

domain_summary
```

The observed matrices go into a `hyperdesign`. The feature matrices stay
outside the fit.

```{r build-hyperdesign}
hd <- hyperdesign(list(
  a = multidesign(observed$a, data.frame(row_id = seq_len(n))),
  b = multidesign(observed$b, data.frame(row_id = seq_len(n))),
  c = multidesign(observed$c, data.frame(row_id = seq_len(n)))
))

hd
```

## What does `feature_correspondences()` mean?

The helper in this vignette has two layers:

1. `feature_correspondence_matrix(Fa, Fb)` builds a sparse matrix whose rows
   index dataset `a` and whose columns index dataset `b`.
2. `feature_correspondences()` converts the nonzero entries of those sparse
   matrices into the edge table expected by `ssma_align()`.

For domains `a` and `b`, the matrix below is `nrow(a) x nrow(b)`. Nonzero entry
`(i, j)` means: row `i` in domain `a` should be treated as semantically close
to row `j` in domain `b`.

```{r sparse-pair-matrix}
W_ab <- feature_correspondence_matrix(
  features_true$a,
  features_true$b,
  top_k = 3,
  mutual = TRUE,
  min_similarity = 0.6
)

dim(W_ab)
Matrix::nnzero(W_ab)
round(Matrix::nnzero(W_ab) / prod(dim(W_ab)), 4)
head(summary(W_ab))

stopifnot(all(dim(W_ab) == c(nrow(observed$a), nrow(observed$b))))
stopifnot(methods::is(W_ab, "sparseMatrix"))
```

Even with only 72 rows per domain, the kept edges are already a small fraction
of the full bipartite matrix. In larger problems you would keep `top_k` small
or build approximate row-neighbour candidates directly, but the geometry is the
same: a sparse **row-by-row** correspondence matrix.

```{r correspondence-table}
corr_full <- feature_correspondences(
  features_true,
  top_k = 3,
  mutual = TRUE,
  min_similarity = 0.6
)

head(corr_full)
table(paste(names(features_true)[corr_full$domain_i],
            names(features_true)[corr_full$domain_j],
            sep = " -> "))

stopifnot(all(corr_full$index_i >= 1L & corr_full$index_i <= n))
stopifnot(all(corr_full$index_j >= 1L & corr_full$index_j <= n))
```

This table is what `ssma_align()` actually consumes. The `weight` column keeps
the cosine similarity, so stronger row-level feature matches get stronger
cross-domain edges.

## End-to-end inference on held-out rows

We will hold out a few rows from domain `a`, rebuild the feature-derived
correspondences on the training rows only, fit SSMA, then project the held-out
rows into the aligned space.

```{r single-split-fit}
fold1 <- multidesign::cv_rows(hd, rows = list(list(a = 1:8)))[[1]]
analysis <- fold1$analysis
assessment <- fold1$assessment

analysis_ids <- lapply(analysis, function(block) block$design$row_id)
features_analysis <- Map(
  function(F, ids) F[ids, , drop = FALSE],
  features_true[names(analysis_ids)],
  analysis_ids
)

corr_analysis <- feature_correspondences(
  features_analysis,
  top_k = 3,
  mutual = TRUE,
  min_similarity = 0.6
)

fit <- ssma_align(
  analysis,
  correspondences = corr_analysis,
  preproc = multivarious::center(),
  ncomp = 2,
  control = ssma_align_control(
    knn = 8,
    rank_per_domain = 12,
    verbose = FALSE
  )
)

nrow(corr_analysis)
stopifnot(nrow(corr_analysis) > 0)
```

Now project the held-out rows from domain `a` with `oos_predict()`, retrieve
their nearest neighbours in the training rows of domains `b` and `c`, and score
those matches in the external feature space.

```{r manual-inference}
query_scores <- oos_predict(fit, assessment$a$x, side = "a")
score_blocks <- score_row_blocks(analysis)

target_scores_b <- fit$s[score_blocks$b, , drop = FALSE]
target_scores_c <- fit$s[score_blocks$c, , drop = FALSE]

nn_b <- latent_knn(query_scores, target_scores_b, k = 3L)
nn_c <- latent_knn(query_scores, target_scores_c, k = 3L)

query_ids <- assessment$a$design$row_id
target_ids_b <- analysis$b$design$row_id
target_ids_c <- analysis$c$design$row_id

feature_sim_b <- row_cosine_similarity(
  features_true$a[query_ids, , drop = FALSE],
  features_true$b[target_ids_b, , drop = FALSE]
)
feature_sim_c <- row_cosine_similarity(
  features_true$a[query_ids, , drop = FALSE],
  features_true$c[target_ids_c, , drop = FALSE]
)

retrieval_summary <- rbind(
  data.frame(
    query_row = query_ids,
    target_domain = "b",
    retrieved_row = target_ids_b[nn_b[, 1]],
    feature_similarity = feature_sim_b[cbind(seq_along(query_ids), nn_b[, 1])],
    stringsAsFactors = FALSE
  ),
  data.frame(
    query_row = query_ids,
    target_domain = "c",
    retrieved_row = target_ids_c[nn_c[, 1]],
    feature_similarity = feature_sim_c[cbind(seq_along(query_ids), nn_c[, 1])],
    stringsAsFactors = FALSE
  )
)

head(retrieval_summary, 8)
aggregate(feature_similarity ~ target_domain, retrieval_summary, mean)

stopifnot(all(is.finite(retrieval_summary$feature_similarity)))
stopifnot(mean(retrieval_summary$feature_similarity) > 0.6)
```

That is the full prediction loop:

1. fit the alignment on observed training matrices;
2. project new rows with `oos_predict()`;
3. retrieve nearby rows in the other domains;
4. ask whether those retrieved rows have similar **row-level features**.

## Cross-validated predictive performance

`cv_alignment_rows()` automates the same logic fold-by-fold. The crucial detail
is that the correspondence table must be rebuilt inside `fit_fn()` from the
training rows only, so held-out rows never leak into the fit.

```{r rowwise-cv}
rows <- list(
  list(a = 1:6),
  list(b = 7:12),
  list(c = 13:18),
  list(a = 19:24),
  list(b = 25:30),
  list(c = 31:36)
)

fit_fn <- function(analysis) {
  analysis_ids <- lapply(analysis, function(block) block$design$row_id)
  feature_blocks <- Map(
    function(F, ids) F[ids, , drop = FALSE],
    features_true[names(analysis_ids)],
    analysis_ids
  )

  corr <- feature_correspondences(
    feature_blocks,
    top_k = 3,
    mutual = TRUE,
    min_similarity = 0.6
  )

  manifoldalign::ssma_align(
    analysis,
    correspondences = corr,
    preproc = multivarious::center(),
    ncomp = 2,
    control = manifoldalign::ssma_align_control(
      knn = 8,
      rank_per_domain = 12,
      verbose = FALSE
    )
  )
}

cv_signal <- manifoldalign::cv_alignment_rows(
  hd,
  rows = rows,
  fit_fn = fit_fn,
  features = features_true,
  k = 3,
  target_pool = "analysis"
)

cv_signal$scores[, c(
  ".fold",
  "mean_top1_similarity",
  "mean_topk_similarity",
  "n_queries",
  "n_pairs"
)]

stopifnot(all(is.finite(cv_signal$scores$mean_top1_similarity)))
stopifnot(all(is.finite(cv_signal$scores$mean_topk_similarity)))
stopifnot(mean(cv_signal$scores$mean_top1_similarity) > 0.7)
```

The score still has the same interpretation as the manual example above. For
each held-out query row, the model projects that row into the aligned space,
retrieves nearby rows from the target domains, and measures the agreement of
their external row-level feature vectors.

## A negative control for the scoring rule

To isolate the evaluation step, we can keep the fitted alignment procedure the
same and shuffle the row-level feature vectors used only for scoring. If the
metric is meaningful, held-out agreement should drop.

```{r negative-control}
set.seed(20260421)
features_shuffled <- lapply(features_true, function(F) {
  F[sample(seq_len(nrow(F))), , drop = FALSE]
})

cv_shuffled <- manifoldalign::cv_alignment_rows(
  hd,
  rows = rows,
  fit_fn = fit_fn,
  features = features_shuffled,
  k = 3,
  target_pool = "analysis"
)

comparison <- rbind(
  aggregate_cv(cv_signal, "Aligned features"),
  aggregate_cv(cv_shuffled, "Shuffled scoring features")
)

comparison

stopifnot(
  mean(cv_signal$scores$mean_top1_similarity) >
    mean(cv_shuffled$scores$mean_top1_similarity) + 0.15
)
```

```{r plot-comparison, echo = FALSE, fig.cap = "Held-out feature agreement drops when the row-level feature vectors are shuffled at scoring time."}
ggplot(comparison, aes(metric, value, fill = scenario)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous(limits = c(-0.1, 1)) +
  labs(x = NULL, y = "Mean feature similarity")
```

## Design takeaway

If external features are only there to define supervision and evaluation, keep
them outside the fit and use them in this order:

1. build sparse row-to-row correspondence matrices between domains;
2. convert those matrices into a correspondence edge table for
   `ssma_align()`;
3. fit on training rows only;
4. project held-out rows with `oos_predict()` and score cross-domain retrieval
   with `cv_alignment_rows()`.

If the external features are themselves a modality you want represented in the
embedding, add them as another domain. That is a different model. In that case
the matching feature rows need to be held out from the fit as well.
