---
title: 'Features, Correspondences, and Predictive Performance'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Features, Correspondences, and Predictive Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(params$family))
}
```

You have multiple data domains, and each row also has an external feature
vector. Those features might come from text embeddings, anatomy, metadata, or
another modality entirely. The alignment problem is:

- use the observed matrices to learn a shared manifold,
- use feature similarity to define cross-domain supervision,
- use held-out feature agreement to define predictive performance.

The important point is that predictive performance is **not** “reconstruct the
feature vector.” It is “does a held-out row land near rows in other domains
whose features are similar?”

## Setup

```{r libraries}
library(manifoldalign)
library(multidesign)
library(ggplot2)
```

```{r helpers, include = FALSE}
feature_correspondences <- function(Fa, Fb) {
  Fa <- as.matrix(Fa)
  Fb <- as.matrix(Fb)

  Fa_norm <- sqrt(rowSums(Fa * Fa))
  Fb_norm <- sqrt(rowSums(Fb * Fb))
  Fa_norm[Fa_norm <= 0] <- 1
  Fb_norm[Fb_norm <= 0] <- 1

  sim <- (Fa / Fa_norm) %*% t(Fb / Fb_norm)
  j_best <- max.col(sim, ties.method = "first")
  i_best <- max.col(t(sim), ties.method = "first")
  keep <- which(i_best[j_best] == seq_along(j_best))

  data.frame(
    domain_i = 1L,
    index_i = keep,
    domain_j = 2L,
    index_j = j_best[keep]
  )
}

aggregate_cv <- function(res, label) {
  data.frame(
    scenario = label,
    metric = c("top-1", "top-k"),
    value = c(
      mean(res$scores$mean_top1_similarity),
      mean(res$scores$mean_topk_similarity)
    )
  )
}
```

## Simulate two aligned domains

We will generate two observed matrices from the same latent variables and keep
separate feature vectors for scoring.

```{r simulate-data}
set.seed(20260420)
n <- 60
latent_dim <- 3

Z <- matrix(rnorm(n * latent_dim), n, latent_dim)
A1 <- matrix(rnorm(latent_dim * 8), latent_dim, 8)
A2 <- matrix(rnorm(latent_dim * 6), latent_dim, 6)

X1 <- Z %*% A1 + matrix(rnorm(n * 8, sd = 0.05), n, 8)
X2 <- Z %*% A2 + matrix(rnorm(n * 6, sd = 0.05), n, 6)

features_true <- list(
  a = Z + matrix(rnorm(n * latent_dim, sd = 0.02), n, latent_dim),
  b = Z + matrix(rnorm(n * latent_dim, sd = 0.02), n, latent_dim)
)
```

The observed data go into a `hyperdesign`. The feature matrices stay outside
the fit and will be used for correspondences and held-out evaluation.

```{r build-hyperdesign}
hd <- hyperdesign(list(
  a = multidesign(X1, data.frame(row_id = seq_len(n))),
  b = multidesign(X2, data.frame(row_id = seq_len(n)))
))

hd
```

## Turn feature similarity into correspondences

Here we use mutual top-1 cosine matches between the feature vectors. In a real
problem you might use a threshold, mutual top-`k`, or weighted anchors.

```{r correspondences}
full_corr <- feature_correspondences(features_true$a, features_true$b)

nrow(full_corr)
head(full_corr)

stopifnot(nrow(full_corr) >= 0.8 * n)
```

That correspondence table is only for the full data. During cross-validation we
must rebuild it on each training split so held-out rows do not leak into the
fit.

## Score held-out rows

`cv_alignment_rows()` does row-wise cross-validation. We hold out rows from
domain `a`, fit on the remaining data, project the held-out rows, and score
them against the training rows in domain `b`.

```{r rowwise-cv}
rows <- list(
  list(a = 1:10),
  list(a = 11:20),
  list(a = 21:30)
)

fit_fn <- function(analysis) {
  ids_a <- analysis$a$design$row_id
  ids_b <- analysis$b$design$row_id

  corr <- feature_correspondences(
    features_true$a[ids_a, , drop = FALSE],
    features_true$b[ids_b, , drop = FALSE]
  )

  manifoldalign::ssma_align(
    analysis,
    correspondences = corr,
    preproc = multivarious::center(),
    ncomp = 2,
    control = manifoldalign::ssma_align_control(
      knn = 8,
      rank_per_domain = 12,
      verbose = FALSE
    )
  )
}

cv_signal <- manifoldalign::cv_alignment_rows(
  hd,
  rows = rows,
  fit_fn = fit_fn,
  features = features_true,
  k = 3,
  target_pool = "analysis"
)

cv_signal$scores

stopifnot(all(is.finite(cv_signal$scores$mean_top1_similarity)))
stopifnot(all(is.finite(cv_signal$scores$mean_topk_similarity)))
stopifnot(mean(cv_signal$scores$mean_top1_similarity) > 0.5)
```

The key score is `mean_top1_similarity`: for each held-out query row, project it
into the fitted space, retrieve its nearest neighbor in the target block, and
measure the cosine similarity of their external feature vectors.

## Does the score detect broken semantic structure?

A good held-out metric should drop when the feature-defined target structure is
destroyed. We can test that by shuffling the target feature rows while keeping
the fitted model fixed.

```{r negative-control}
set.seed(20260421)
features_shuffled <- list(
  a = features_true$a,
  b = features_true$b[sample(seq_len(n)), , drop = FALSE]
)

cv_shuffled <- manifoldalign::cv_alignment_rows(
  hd,
  rows = rows,
  fit_fn = fit_fn,
  features = features_shuffled,
  k = 3,
  target_pool = "analysis"
)

comparison <- rbind(
  aggregate_cv(cv_signal, "Aligned features"),
  aggregate_cv(cv_shuffled, "Shuffled target features")
)

comparison

stopifnot(
  mean(cv_signal$scores$mean_top1_similarity) >
    mean(cv_shuffled$scores$mean_top1_similarity) + 0.2
)
```

```{r plot-comparison, echo = FALSE, fig.cap = "Held-out feature agreement drops when the target feature rows are shuffled, which is what we want from a predictive scoring rule."}
ggplot(comparison, aes(metric, value, fill = scenario)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous(limits = c(-0.1, 1)) +
  labs(x = NULL, y = "Mean feature similarity")
```

This is the predictive story:

- the alignment is fit on the observed matrices,
- feature similarity defines the semantic notion of a good cross-domain match,
- held-out performance asks whether new rows land near semantically similar
  rows in the fitted space.

## Design takeaway

If your features are only there to define supervision and evaluation, keep
them **outside** the fit and use them to:

1. build correspondence edges on the training rows,
2. score held-out rows with `cv_alignment_rows()`.

If the features are a real modality you want represented in the embedding, you
can also add them as another domain in the alignment itself. That is a
different model, and cross-validation then has to hold out the matching feature
rows too.
