---
title: Generalized PCA Alignment with gpca_align
output: rmarkdown::html_vignette
vignette: '%\VignetteIndexEntry{Generalized PCA Alignment with gpca_align} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}'
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`gpca_align()` implements the Generalized PCA alignment described in
`R/gpca_align.R`.  It constructs a coupled metric across domains and then solves
a generalized PCA problem to obtain a shared latent embedding.  The helper
`gpca_align_control()` exposes safeguards for graph sparsification, label
balancing and dense-matrix memory limits, making it practical on real data.

This vignette walks through a minimal pipeline:

1. Load a reusable **hyperdesign** dataset with three related domains.
2. Run `gpca_align()` and inspect the aligned scores.
3. Tune graph sparsity and balancing with `gpca_align_control()`.

The code below only relies on public package APIs, so you can copy–paste it when
exploring your own data.

## Setup

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
library(manifoldalign)
library(multidesign)
library(multivarious)
library(tibble)
library(dplyr)
library(ggplot2)
library(purrr)
```

## Constructing a Hyperdesign

We reuse the packaged benchmark dataset so every alignment vignette operates on
identical inputs.  It contains three domains with two latent classes and
moderate domain-specific transformations.

```{r synthetic-data}
alignment_benchmark <- manifoldalign::alignment_benchmark
domain_list <- lapply(alignment_benchmark$domains, function(dom) {
  multidesign(dom$x, dom$design)
})
hd <- hyperdesign(domain_list)
labels <- alignment_benchmark$labels
domain_names <- names(domain_list)
domain_sizes <- vapply(domain_list, function(dom) nrow(dom$x), integer(1))
```

```{r gpca-helpers}
# Standardise columns while guarding against zero variance
zscore_columns <- function(mat) {
  center <- sweep(mat, 2, colMeans(mat), "-")
  sds <- sqrt(colMeans(center^2))
  sds[sds == 0] <- 1
  center / sds
}
```

When `data` is a `hyperdesign`, call `gpca_align()` and supply the label column
name via the `y` argument.

```{r run-gpca-align}
base_preproc <- replicate(length(domain_names), multivarious::pass(), simplify = FALSE)

gpca_fit <- gpca_align(
  hd,
  y = condition,
  ncomp = 2,
  u = 0.6,           # balance within/between alignment
  lambda = 1e-2,     # light ridge regularisation
  preproc = base_preproc
)

str(gpca_fit, max.level = 1)
```

```{r gpca-metrics}
# RMS alignment between domain pairs
rms_alignment <- function(scores, sizes, names) {
  offsets <- c(0, cumsum(sizes))
  idx <- setNames(vector('list', length(names)), names)
  for (i in seq_along(names)) {
    idx[[names[i]]] <- seq(offsets[i] + 1L, offsets[i + 1L])
  }
  purrr::map_dfr(combn(names, 2, simplify = FALSE), function(pair) {
    i_idx <- idx[[pair[1]]]
    j_idx <- idx[[pair[2]]]
    mat_i <- scores[i_idx, , drop = FALSE]
    mat_j <- scores[j_idx, , drop = FALSE]
    tibble(
      domain_i = pair[1],
      domain_j = pair[2],
      rms = sqrt(mean(rowSums((mat_i - mat_j)^2)))
    )
  })
}

rms_alignment(as.matrix(gpca_fit$s), domain_sizes, domain_names)
```

The result is a `multiblock_biprojector`, so we can extract scores and feature
loadings just like other `multivarious` models. For plotting below we z-score the
aligned scores so that the scale matches other vignettes.

You can optionally pass a domain-specific preprocessing list (e.g.
`list(center(), center(), center())`) if you want to centre or scale each block
before alignment.  Passing the `base_preproc` list (built above with
`multivarious::pass()`) keeps preprocessing as the identity—which keeps this
example lightweight while still supporting different feature dimensions.

```{r inspect-scores}
score_tbl <- as_tibble(zscore_columns(as.matrix(gpca_fit$s)), .name_repair = "minimal")
colnames(score_tbl) <- paste0("comp", seq_len(ncol(score_tbl)))

scores <- score_tbl %>%
  mutate(
    sample = seq_len(nrow(.)),
    domain = rep(domain_names, times = domain_sizes),
    condition = rep(labels, length(domain_names))
  )

head(scores)
```

### Visualising the Shared Embedding

```{r plot-scores, fig.width = 6, fig.height = 4}
 ggplot(scores, aes(x = comp1, y = comp2, colour = condition, shape = domain)) +
  geom_point(size = 2.2, alpha = 0.85) +
  labs(
    title = "GPCA alignment: shared 2D embedding",
    x = "Component 1",
    y = "Component 2"
  ) +
  theme_minimal()
```

## Controlling Graph Sparsification and Balancing

The optional `control` argument feeds through to
`gpca_align_control()`, letting you adjust k-NN sparsification, label balancing
and memory caps.  Here we build a denser within-domain graph but down-weight the
majority class via label balancing.

```{r control-example}
ctrl <- gpca_align_control(
  knn = 10,
  knn_mode = "mutual",
  balance = "within",
  balance_power = 0.5,
  normalize = "edges",
  verbose = FALSE
)

gpca_dense <- gpca_align(
  hd,
  y = condition,
  ncomp = 2,
  control = ctrl,
  preproc = base_preproc
)

summary(gpca_dense$sdev)
```

When working with larger domains, consider leaving `knn = NA` (the default) to
use fully dense metrics or lower the `max_dense_elems` parameter to enforce a
safe memory ceiling.  If `gpca_align()` detects that densifying the block matrix
would exceed the limit, it aborts with a diagnostic message so you can tighten
the sparsification or dimensionality reduction beforehand.

## Next Steps

- `gpca_align()` returns a standard `multivarious` projector, so you can call
  `predict()` on new samples (via `multivarious::multiblock_biprojector`).
- The RMS alignment summary above shows how closely each domain pair agrees in the
  coupled space; values near zero indicate strong agreement.
- Use `gpca_align_control()` to add sparsification or label balancing when operating
  at larger scale.
- Pair `gpca_align()` with `kema()` or `grasp()` in the same pipeline to compare
  linear vs kernel alignment strategies.

This vignette is intentionally lightweight—swap in your own domains or expand
chunks to benchmark larger settings.
