---
title: "Spectral Graph Alignment with grasp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spectral Graph Alignment with grasp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`grasp()` aligns two graph domains by computing spectral embeddings, building
multi-scale diffusion descriptors, and solving for a functional map plus a node
assignment.  The method is unsupervised—the latent classes are only consulted in
this vignette for diagnostic plots.

We work with the two graph domains provided by `alignment_benchmark` so the
results are directly comparable to the `cone_align`, `parrot`, and `kema`
vignettes.

## Setup

```{r setup}
library(manifoldalign)
library(multidesign)
library(multivarious)
library(tibble)
library(dplyr)
library(ggplot2)
library(scales)
```

## Loading the Benchmark Pair

```{r data}
alignment_benchmark <- manifoldalign::alignment_benchmark
pair_domains <- alignment_benchmark$domains[1:2]

pair_multidesign <- lapply(pair_domains, function(dom) {
  multidesign::multidesign(dom$x, dom$design)
})

pair_names <- names(pair_multidesign)
node_counts <- vapply(pair_multidesign, function(dom) nrow(dom$x), integer(1))

hd_pair <- multidesign::hyperdesign(pair_multidesign)
```

## Running `grasp`

```{r run-grasp}
grasp_internal <- grasp(
  hd_pair,
  ncomp = 20,
  q_descriptors = 80,
  sigma = 0.8,
  lambda = 0.1,
  solver = "linear"
)

str(grasp_internal$assignment)
```

### Assignment Diagnostics

The `assignment` component maps nodes in the second domain to nodes in the first
(reference) domain.  We compare the assigned pairs with the underlying class
labels to quantify alignment quality—remember that the optimiser itself never
sees these labels.

```{r grasp-diagnostics}
assignment <- grasp_internal$assignment
ref_design <- pair_multidesign[[1]]$design
cmp_design <- pair_multidesign[[2]]$design

class_accuracy <- mean(ref_design$condition == cmp_design$condition[assignment])
identity_fraction <- mean(assignment == seq_along(assignment))

tibble(
  metric = c("Class agreement", "Exact identity"),
  value = c(class_accuracy, identity_fraction)
)
```

```{r grasp-error}
n_ref <- node_counts[1]
embed1 <- grasp_internal$s[seq_len(n_ref), , drop = FALSE]
embed2 <- grasp_internal$s[n_ref + seq_len(n_ref), , drop = FALSE]
final_err <- sqrt(mean(rowSums((embed1 - embed2)^2)))

tibble(
  metric = "Post-alignment RMS error",
  value = final_err
)
```

### Visualising the Aligned Embeddings

We stack the aligned embeddings for both domains (as returned by `grasp()`) and
colour points by their latent class.

```{r grasp-plot, fig.width = 6, fig.height = 4}
embed1 <- grasp_internal$s[seq_len(n_ref), , drop = FALSE]
embed2 <- grasp_internal$s[n_ref + seq_len(n_ref), , drop = FALSE]

score_tbl <- as_tibble(rbind(embed1, embed2), .name_repair = "minimal")
colnames(score_tbl) <- paste0("comp", seq_len(ncol(score_tbl)))

scores <- score_tbl %>%
  mutate(
    sample = seq_len(nrow(.)),
    domain = rep(pair_names, each = nrow(embed1)),
    condition = c(ref_design$condition, cmp_design$condition[assignment])
  )

 ggplot(scores, aes(x = comp1, y = comp2, colour = condition, shape = domain)) +
  geom_point(size = 2, alpha = 0.85) +
  labs(title = "GRASP alignment: first two components",
       x = "Component 1", y = "Component 2") +
  theme_minimal()
```

## Summary

- `grasp()` performs spectral alignment without supervision; class labels are
  only used here for evaluation.
- The post-alignment RMS error can be used as a quick sanity check that the
  returned embeddings are in a comparable coordinate system.
- Because all alignment vignettes reuse the same `alignment_benchmark` domains,
  you can compare GRASP, CONE-Align, PARROT, GPCA, and KEMA directly.
