---
title: Kernel Manifold Alignment with kema
output: rmarkdown::html_vignette
vignette: '%\VignetteIndexEntry{Kernel Manifold Alignment with kema} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}'
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(params$family))
}
```

## Overview

`kema()` aligns multiple domains by projecting them into a shared latent space
where both manifold geometry and class structure are preserved.  The function
supports:

- **Full KEMA** (exact or regression solver) on the complete kernel matrices
- **REKEMA** landmark approximations via `sample_frac`
- A scalable path (enabled with `options(kema.scalable_mode = TRUE)`) that
  replaces dense label graphs and block-diagonal kernels by compact operators

This vignette demonstrates each mode on the shared
`alignment_benchmark` dataset so that comparisons across alignment methods are
directly comparable.

## Setup

```{r setup}
library(manifoldalign)
library(multidesign)
library(multivarious)
library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
```

## Benchmark Hyperdesign

We load `alignment_benchmark`, which provides three domains with two latent
classes and modest domain-specific transformations.

```{r data}
alignment_benchmark <- manifoldalign::alignment_benchmark
domain_list <- lapply(alignment_benchmark$domains, function(dom) {
  multidesign(dom$x, dom$design)
})
hd <- hyperdesign(domain_list)
labels <- alignment_benchmark$labels
domain_names <- names(domain_list)
domain_sizes <- vapply(domain_list, function(dom) nrow(dom$x), integer(1))
replicated_labels <- rep(labels, length(domain_names))
```

```{r kema-helpers, include = FALSE}
zscore_columns <- function(mat) {
  center <- sweep(mat, 2, colMeans(mat), "-")
  sds <- sqrt(colMeans(center^2))
  sds[sds == 0] <- 1
  center / sds
}

rms_alignment <- function(scores, sizes, group_names) {
  offsets <- c(0, cumsum(sizes))
  idx <- setNames(vector("list", length(group_names)), group_names)
  for (i in seq_along(group_names)) {
    idx[[group_names[i]]] <- seq(offsets[i] + 1L, offsets[i + 1L])
  }
  purrr::map_dfr(combn(group_names, 2, simplify = FALSE), function(pair) {
    mat_i <- scores[idx[[pair[1]]], , drop = FALSE]
    mat_j <- scores[idx[[pair[2]]], , drop = FALSE]
    tibble(domain_i = pair[1], domain_j = pair[2],
           rms = sqrt(mean(rowSums((mat_i - mat_j)^2))))
  })
}
```

## Full KEMA (Regression Solver)

The regression solver is the default.  It solves an eigenproblem on graph
Laplacians and follows with a ridge-regression step to recover kernel
coefficients.  Because the domains have different feature dimensions we use
`multivarious::pass()` (no preprocessing) to avoid broadcasting a shared
preprocessor across incompatible shapes.

```{r kema-regression}
reg_fit <- kema(
  hd,
  y = condition,
  ncomp = 2,
  knn = 3,
  sigma = 0.8,
  solver = "regression",
  preproc = multivarious::pass()
)

str(reg_fit, max.level = 1)
```

```{r kema-reg-metrics}
rms_alignment(as.matrix(reg_fit$s), domain_sizes, domain_names)
```

We plot z-scored aligned scores (each component standardized across all domains);
this keeps the visual scale comparable while preserving class separation.

```{r reg-plot, fig.width = 6, fig.height = 4}
score_tbl <- as_tibble(zscore_columns(as.matrix(reg_fit$s)), .name_repair = "minimal")
colnames(score_tbl) <- paste0("comp", seq_len(ncol(score_tbl)))

reg_scores <- score_tbl %>%
  mutate(
    domain = rep(domain_names, times = domain_sizes),
    condition = replicated_labels
  )

ggplot(reg_scores, aes(x = comp1, y = comp2, colour = condition, shape = domain)) +
  geom_point(size = 2.2, alpha = 0.85) +
  labs(title = "KEMA regression solver", x = "Component 1", y = "Component 2") +
  theme_minimal()
```

## Full KEMA (Exact Solver)

The exact solver tackles the generalized eigenvalue problem directly.  It is
more faithful but slower on very large kernels.

```{r kema-exact}
exact_fit <- kema(
  hd,
  y = condition,
  ncomp = 2,
  knn = 3,
  sigma = 0.8,
  solver = "exact",
  preproc = multivarious::pass()
)

exact_fit$eigenvalues$values
```

```{r kema-exact-metrics}
rms_alignment(as.matrix(exact_fit$s), domain_sizes, domain_names)
```

A quick sanity check compares the subspaces recovered by the two solvers.

```{r compare-solvers}
subspace_similarity <- function(A, B) {
  A <- as.matrix(A)
  B <- as.matrix(B)
  Qa <- qr.Q(qr(A))
  Qb <- qr.Q(qr(B))
  svd(t(Qa) %*% Qb, nu = 0, nv = 0)$d
}

subspace_similarity(reg_fit$s[, 1:2], exact_fit$s[, 1:2])
```

## REKEMA (Landmark Approximation)

Sampling landmarks via `sample_frac` reduces the kernel size.  Here we keep 50%
of the samples per domain and run the exact solver on the reduced system.

```{r rekema}
rekema_fit <- kema(
  hd,
  y = condition,
  ncomp = 2,
  knn = 3,
  sigma = 0.8,
  solver = "exact",
  sample_frac = 0.5,
  preproc = multivarious::pass()
)

rekema_fit$eigenvalues$solver
```

```{r kema-rekema-metrics}
rms_alignment(as.matrix(rekema_fit$s), domain_sizes, domain_names)
```

Even with landmarks the subspace remains close to the full exact solution.

```{r compare-rekema}
subspace_similarity(exact_fit$s[, 1:2], rekema_fit$s[, 1:2])
```

## Scalable Mode (Operator-Based)

Enabling `kema.scalable_mode` activates the low-rank label-factor and
matrix-free kernel path introduced in the accompanying refactor.  This keeps the
API identical while avoiding dense `n \times n` intermediates.

```{r scalable-mode}
old_opts <- options(kema.scalable_mode = TRUE)
on.exit(options(old_opts), add = TRUE)

scalable_fit <- kema(
  hd,
  y = condition,
  ncomp = 2,
  knn = 3,
  sigma = 0.8,
  solver = "exact",
  preproc = multivarious::pass()
)

scalable_fit$eigenvalues$solver
```

```{r kema-scalable-metrics}
rms_alignment(as.matrix(scalable_fit$s), domain_sizes, domain_names)
```

The scalable path should agree closely with the baseline exact solution.

```{r compare-scalable}
subspace_similarity(exact_fit$s[, 1:2], scalable_fit$s[, 1:2])
```

## Tuning Key Parameters

The graph neighbourhood size `knn`, the Gaussian kernel width `sigma`, and the trade-off `u` are the most immediate levers in practice. Smaller `sigma` emphasises local structure, while larger `u` leans on label agreement. The grid below reports the mean RMS alignment (smaller is better) for a few representative settings.

```{r kema-parameter-grid}
parameter_grid <- tidyr::expand_grid(
  sigma = c(0.5, 0.8, 1.1),
  knn = c(3, 6),
  u = c(0.35, 0.5, 0.7)
)

grid_summary <- parameter_grid %>%
  mutate(
    mean_rms = purrr::pmap_dbl(
      list(sigma, knn, u),
      function(sigma, knn, u) {
        fit <- kema(
          hd,
          y = condition,
          ncomp = 2,
          knn = knn,
          sigma = sigma,
          u = u,
          solver = "regression",
          preproc = multivarious::pass()
        )

        mean(rms_alignment(as.matrix(fit$s), domain_sizes, domain_names)$rms)
      }
    )
  ) %>%
  arrange(mean_rms)

grid_summary
```

The embedding geometry shifts noticeably between configurations with the lowest and highest mean RMS:

```{r kema-parameter-embeddings, fig.width = 6, fig.height = 4}
best_cfg <- grid_summary %>% slice_min(mean_rms, n = 1, with_ties = FALSE)
worst_cfg <- grid_summary %>% slice_max(mean_rms, n = 1, with_ties = FALSE)

collect_scores <- function(cfg, tag) {
  fit <- kema(
    hd,
    y = condition,
    ncomp = 2,
    knn = cfg$knn[[1]],
    sigma = cfg$sigma[[1]],
    u = cfg$u[[1]],
    solver = "regression",
    preproc = multivarious::pass()
  )

  tbl <- as_tibble(zscore_columns(as.matrix(fit$s)), .name_repair = "minimal")
  colnames(tbl) <- paste0("comp", seq_len(ncol(tbl)))

  tbl %>%
    mutate(
      domain = rep(domain_names, times = domain_sizes),
      condition = replicated_labels,
      config = tag
    )
}

comparison_scores <- dplyr::bind_rows(
  collect_scores(best_cfg, "low mean RMS"),
  collect_scores(worst_cfg, "high mean RMS")
)

 ggplot(comparison_scores, aes(x = comp1, y = comp2, colour = condition, shape = domain)) +
  geom_point(alpha = 0.85, size = 2.1) +
  facet_wrap(~ config) +
  labs(
    title = "Effect of kernel and graph parameters",
    x = "Component 1",
    y = "Component 2"
  ) +
  theme_minimal()
```

In general, tighter kernels (`sigma = 0.5`) with modest neighbourhoods (`knn = 3`) and balanced supervision (`u = 0.5`) yield the most consistent alignment on this benchmark, while overly wide kernels and heavy supervision degrade overlap. Use these diagnostics as a template when tuning on larger data.


## Summary

- Use the **regression solver** for quick prototypes; switch to **exact** if the
  approximation quality is insufficient.
- **REKEMA** (`sample_frac < 1`) reduces the computational burden when full
  kernels are too large.
- Toggle `options(kema.scalable_mode = TRUE)` to engage the new low-rank
  operator path.
- The RMS alignment tables highlight how closely each solver agrees across domains;
  lower values correspond to tighter alignment.

These examples are deliberately small for clarity—swap in your own domains and
adjust `sigma`, `knn`, or control options to match real data.
