---
title: Low-Rank Alignment with lowrank_align
output: rmarkdown::html_vignette
vignette: '%\VignetteIndexEntry{Low-Rank Alignment with lowrank_align} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}'
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(params$family))
}

rms_alignment <- function(scores, sizes, names) {
  offsets <- c(0, cumsum(sizes))
  idx <- setNames(vector("list", length(names)), names)
  for (i in seq_along(names)) {
    idx[[names[i]]] <- seq(offsets[i] + 1L, offsets[i + 1L])
  }
  purrr::map_dfr(combn(names, 2, simplify = FALSE), function(pair) {
    mat_i <- scores[idx[[pair[1]]], , drop = FALSE]
    mat_j <- scores[idx[[pair[2]]], , drop = FALSE]
    tibble(domain_i = pair[1], domain_j = pair[2],
           rms = sqrt(mean(rowSums((mat_i - mat_j)^2))))
  })
}
```

## Overview

`lowrank_align()` balances low-rank reconstruction with
similarity-derived Laplacian structure. The method was designed for partially
observed labels: unlabeled samples contribute to the low-rank term while labeled
samples supply supervision through a similarity kernel. This vignette walks
through a semi-supervised alignment workflow:

1. Build a three-domain `hyperdesign` with missing labels in two domains.
2. Construct a label-aware similarity function for the Laplacian term.
3. Fit `lowrank_align()` and inspect the shared embedding.
4. Summarise alignment quality with RMS discrepancies between domains.

All code uses public APIs so you can transplant the pattern to your own data.

## Setup

```{r setup}
library(manifoldalign)
library(multidesign)
library(multivarious)
library(tibble)
library(dplyr)
library(ggplot2)
library(purrr)
```

## Building a Partially-Labeled Hyperdesign

We start from the packaged `alignment_benchmark` dataset, convert each domain to
`multidesign` objects, and deliberately mask a subset of labels to illustrate the
semi-supervised behaviour.

```{r hyperdesign}
alignment_benchmark <- manifoldalign::alignment_benchmark

# Convert each domain to a multidesign object
domain_list <- lapply(alignment_benchmark$domains, function(dom) {
  multidesign(dom$x, dom$design)
})

domain_names <- names(domain_list)
domain_sizes <- vapply(domain_list, function(dom) nrow(dom$x), integer(1))

# Hide labels in alternating batches for the first two domains
semi_domain_list <- domain_list
semi_domain_list[[1]]$design$condition[seq(1, domain_sizes[1], by = 4)] <- NA
semi_domain_list[[2]]$design$condition[seq(2, domain_sizes[2], by = 4)] <- NA

semi_hd <- hyperdesign(semi_domain_list)
observed_labels <- purrr::map(semi_domain_list, ~ .x$design$condition) %>% unlist()

label_status <- tibble(
  domain = rep(domain_names, times = domain_sizes),
  status = if_else(is.na(observed_labels), "unlabeled", "labeled")
) %>%
  count(domain, status)

label_status
```

## Similarity Function for the Laplacian Term

`lowrank_align` expects a `simfun` that turns the (possibly NA) label vector into
an affinity matrix. The helper below links samples that share the same observed
label and leaves unlabeled rows/columns at zero, which dovetails with the
Laplacian masking performed inside `lowrank_align()`.

```{r similarity}
label_similarity <- function(labels) {
  labs <- as.character(labels)
  valid_levels <- unique(labs[!is.na(labs)])
  n <- length(labs)
  sim <- matrix(0, nrow = n, ncol = n)

  for (lvl in valid_levels) {
    idx <- which(labs == lvl)
    if (length(idx) > 1) {
      sim[idx, idx] <- 1
    }
  }

  diag(sim) <- 0
  sim
}

# Quick sanity check on a small slice
label_similarity(observed_labels[1:6])
```

## Fitting `lowrank_align`

We request three components with a moderate balance (`mu = 0.35`) between the
low-rank term `M` and the Laplacian term `L`. The operator-based solver keeps the
computation sparse without forming dense matrices.

```{r fit-lowrank}
lowrank_fit <- lowrank_align(
  semi_hd,
  y = condition,
  simfun = label_similarity,
  mu = 0.35,
  ncomp = 3,
  sv_thresh = 0.5,
  solver = "operator"
)

str(lowrank_fit, max.level = 1)
lowrank_fit$sdev
```

## Inspecting the Shared Embedding

The returned object is a `multiblock_biprojector`, so we can extract aligned
scores and SQL-like metadata just like other `multivarious` models.

```{r scores}
scores_tbl <- as_tibble(as.matrix(lowrank_fit$s), .name_repair = "minimal")
colnames(scores_tbl) <- paste0("comp", seq_len(ncol(scores_tbl)))

true_labels <- rep(alignment_benchmark$labels, times = length(domain_names))

scores <- scores_tbl %>%
  mutate(
    sample = row_number(),
    domain = rep(domain_names, times = domain_sizes),
    observed_condition = observed_labels,
    condition = as.character(true_labels),
    label_status = if_else(is.na(observed_condition), "unlabeled", "labeled"),
    alpha = if_else(label_status == "labeled", 0.95, 0.35)
  )

head(scores)
```

```{r plot-scores, fig.width = 6, fig.height = 4}
 ggplot(scores, aes(x = comp1, y = comp2, colour = condition, shape = domain)) +
  geom_point(aes(alpha = alpha), size = 2.1) +
  scale_alpha_identity() +
  labs(
    title = "Low-rank alignment: shared embedding",
    x = "Component 1",
    y = "Component 2"
  ) +
  theme_minimal()
```

## Domain Agreement Diagnostics

Following the other alignment vignettes, RMS discrepancies quantify how tightly
the domains co-register in the shared space. Values near zero indicate strong
agreement between domain pairs.

```{r rms}
rms_alignment(as.matrix(lowrank_fit$s), domain_sizes, domain_names)
```

## Summary

- `lowrank_align()` naturally accommodates missing labels: unlabeled
  nodes stay in the low-rank term while the Laplacian only couples labeled pairs.
- The label-aware similarity function above produces a sparse Laplacian without
  touching unobserved samples, matching the algorithm's expectations.
- On the benchmark data the first three components cleanly separate the two
  latent classes and keep per-domain RMS discrepancies low despite masked
  supervision.
- Switch to the explicit solver or adjust `sv_thresh` when working with smaller
  problems where dense matrices are affordable.

```{r, include=FALSE}
# Ensure chunks do not leak large objects when vignette is built
rm(alignment_benchmark, domain_list, semi_domain_list, semi_hd, lowrank_fit)
```
