---
title: Multiset Manifold Alignment with mma_align_multiple
output: rmarkdown::html_vignette
vignette: '%\VignetteIndexEntry{Multiset Manifold Alignment with mma_align_multiple}
  %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}'
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dev = "png"
)
```

## Overview

`mma_align_multiple()` aligns three or more domains by combining heat-kernel
spectral embeddings, eigensignature-based eigenbasis registration, and an
EM refinement step that supports both reference matching and joint consensus
templates.  The function exposes controls for histogram versus Wasserstein
signatures, gating heuristics for repeated eigenvalues, and final Hungarian
assignments when hard correspondences are needed.

This vignette follows the style of the other tutorials in the package: we build a
multiset from the shared `alignment_benchmark`, run both reference and consensus
modes, and inspect diagnostics returned by `alignment_quality()`.

## Setup

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
library(manifoldalign)
library(multidesign)
library(tibble)
library(dplyr)
library(ggplot2)
```

## Preparing the Multiset Example

```{r prepare-multiset}
alignment_benchmark <- manifoldalign::alignment_benchmark
raw_domains <- alignment_benchmark$domains

multi_domains <- lapply(raw_domains, function(dom) {
  multidesign::multidesign(dom$x, dom$design)
})

multi_names <- names(multi_domains)
node_counts <- vapply(multi_domains, function(dom) nrow(dom$x), integer(1))

hd_multi <- multidesign::hyperdesign(multi_domains)

tibble(
  domain = multi_names,
  nodes = node_counts,
  features = vapply(multi_domains, function(dom) ncol(dom$x), integer(1))
)
```

The latent classes shipped with the benchmark are used only for diagnostics.

```{r label-distribution}
table(multi_domains[[1]]$design$condition)
```

## Reference Mode Alignment

We first align every domain to `domain1` with eigensignature gating, hypersphere
normalisation, and a final Hungarian pass to extract hard correspondences.

```{r run-reference}
set.seed(1)
mma_ref <- mma_align_multiple(
  hd_multi,
  ref_idx = 1,
  ncomp = 8,
  sigma = 0.7,
  embedding = "ctd",
  normalize = "hypersphere",
  match_to = "reference",
  signature_method = "hybrid",
  signature_gate_multiplicity = TRUE,
  signature_retry_if_weak = TRUE,
  final_assignment = "hungarian"
)

str(mma_ref, max.level = 1)
```

```{r quality-reference}
quality_ref <- alignment_quality(mma_ref)

tibble::as_tibble(quality_ref$per_domain)

tibble::as_tibble(quality_ref$global)
```

The per-domain diagnostics expose EM convergence, average eigensignature
confidence, and MSE under both hard and soft correspondences.  We can relate the
assignments back to the hidden labels of the benchmark.

```{r reference-accuracy}
ref_conditions <- multi_domains[[1]]$design$condition

class_accuracy <- vapply(seq_along(multi_names)[-1], function(idx) {
  perm <- mma_ref$assignment[[idx]]
  keep <- which(!is.na(perm))
  if (!length(keep)) return(NA_real_)
  mean(ref_conditions[perm[keep]] == multi_domains[[idx]]$design$condition[keep])
}, numeric(1))

coverage <- quality_ref$per_domain$coverage[-1]

accuracy_tbl <- tibble(
  domain = multi_names[-1],
  class_accuracy = class_accuracy,
  hard_match_coverage = coverage
)

accuracy_tbl
```

The aligned embeddings combine all domains row-wise.  Visual inspection of the
first two components highlights the shared latent structure recovered by
`mma_align_multiple()`.

```{r reference-plot, fig.alt = "Scatter of MMA reference-mode embeddings across domains and latent classes"}
score_tbl <- as_tibble(as.matrix(mma_ref$s), .name_repair = "minimal")
colnames(score_tbl) <- paste0("comp", seq_len(ncol(score_tbl)))

score_tbl <- score_tbl %>%
  mutate(
    domain = rep(multi_names, times = node_counts),
    condition = unname(unlist(lapply(multi_domains, function(dom) dom$design$condition)))
  )

 ggplot(score_tbl, aes(x = comp1, y = comp2, colour = condition, shape = domain)) +
  geom_point(alpha = 0.85, size = 2) +
  labs(title = "Reference-mode MMA embeddings", x = "Component 1", y = "Component 2") +
  theme_minimal()
```

## Consensus Template Mode

Consensus mode learns a shared orthogonal template while allowing uneven domain
sizes.  We down-sample `domain3` to 60 nodes to illustrate the automatic choice
of consensus centres (`min` by default).

```{r prepare-consensus}
set.seed(42)
consensus_domains <- raw_domains
keep_idx <- sort(sample(seq_len(nrow(consensus_domains[[3]]$x)), 60))
consensus_domains[[3]]$x <- consensus_domains[[3]]$x[keep_idx, , drop = FALSE]
consensus_domains[[3]]$design <- consensus_domains[[3]]$design[keep_idx, ]

multi_consensus <- lapply(consensus_domains, function(dom) {
  multidesign::multidesign(dom$x, dom$design)
})

consensus_names <- names(multi_consensus)
consensus_counts <- vapply(multi_consensus, function(dom) nrow(dom$x), integer(1))

hd_consensus <- multidesign::hyperdesign(multi_consensus)

tibble(
  domain = consensus_names,
  nodes = consensus_counts
)
```

```{r run-consensus}
set.seed(99)
mma_consensus <- mma_align_multiple(
  hd_consensus,
  ref_idx = 1,
  ncomp = 8,
  sigma = 0.7,
  embedding = "ctd",
  normalize = "hypersphere",
  match_to = "consensus",
  consensus_centers = "min",
  consensus_init = "ref",
  signature_method = "w2",
  eig_penalty = 0.01,
  signature_gate_multiplicity = TRUE,
  final_assignment = "hungarian",
  em_outlier = 0.05
)

list(
  consensus_centers = nrow(mma_consensus$consensus),
  em_iterations = mma_consensus$em_stats$iterations,
  converged = mma_consensus$em_stats$converged
)
```

```{r quality-consensus}
quality_consensus <- alignment_quality(mma_consensus)

tibble::as_tibble(quality_consensus$per_domain)

tibble::as_tibble(quality_consensus$global)
```

The consensus template captures the shared geometry in a compact basis.  Each
domain's embeddings sit close to the learned centres even though `domain3` has
fewer nodes.

```{r consensus-plot, fig.alt = "Consensus-mode MMA embeddings with domain points and learned centres"}
consensus_scores <- as_tibble(as.matrix(mma_consensus$s), .name_repair = "minimal")
colnames(consensus_scores) <- paste0("comp", seq_len(ncol(consensus_scores)))

consensus_scores <- consensus_scores %>%
  mutate(
    domain = rep(consensus_names, times = consensus_counts),
    condition = unname(unlist(lapply(multi_consensus, function(dom) dom$design$condition)))
  )

centers_tbl <- tibble(
  comp1 = mma_consensus$consensus[, 1],
  comp2 = mma_consensus$consensus[, 2],
  domain = "consensus",
  condition = factor("template")
)

plot_tbl <- bind_rows(
  mutate(consensus_scores, marker = "nodes"),
  mutate(centers_tbl, marker = "centres")
)

 ggplot(plot_tbl, aes(x = comp1, y = comp2, colour = condition, shape = marker)) +
  geom_point(alpha = 0.85, size = 2) +
  labs(title = "Consensus-mode MMA embeddings", x = "Component 1", y = "Component 2") +
  theme_minimal()
```

Consensus assignments point each domain back to indices of the template.  The
coverage column in `alignment_quality()` indicates how many nodes land on a
unique centre; increasing `em_outlier` or switching `final_assignment` to
`"argmax"` can trade precision for recall when outliers are expected.

## Summary

- `mma_align_multiple()` unifies eigensignature alignment and EM refinement for both reference and consensus workflows.
- Histogram/Wasserstein signatures with gating stabilise eigenbasis matching before the EM loop.
- `alignment_quality()` reports convergence, coverage, and confidence diagnostics without relying on labels.
- Consensus mode automatically adapts the template size when domains have different node counts.
- Hard assignments remain optional; the EM posteriors can drive downstream soft analyses when coverage is low.
