---
title: Position-Aware OT Alignment with parrot
output: rmarkdown::html_vignette
vignette: '%\VignetteIndexEntry{Position-Aware OT Alignment with parrot} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}'
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`parrot()` aligns two networks using position-aware optimal transport.  It builds
Random-Walk-with-Restart (RWR) descriptors for each domain, combines them with
feature distances, and solves an entropy-regularised transport problem that can
optionally incorporate known anchor correspondences.

We demonstrate the workflow on the shared `alignment_benchmark` dataset, adding a
small set of anchors to guide the semi-supervised transport.

## Setup

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
library(manifoldalign)
library(multidesign)
library(multivarious)
library(tibble)
library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)
```

## Preparing Anchored Domains

```{r data}
alignment_benchmark <- manifoldalign::alignment_benchmark
pair_domains <- alignment_benchmark$domains[1:2]

# Convert to multidesign objects
pair_multidesign <- lapply(pair_domains, function(dom) {
  multidesign::multidesign(dom$x, dom$design)
})

# Add matching anchors (10 nodes) to each domain's design frame
set.seed(123)
num_anchors <- 10
anchor_idx <- sample(seq_len(nrow(pair_multidesign[[1]]$x)), num_anchors)

pair_multidesign[[1]]$design$anchor_id <- NA_integer_
pair_multidesign[[1]]$design$anchor_id[anchor_idx] <- anchor_idx
pair_multidesign[[2]]$design$anchor_id <- NA_integer_
pair_multidesign[[2]]$design$anchor_id[anchor_idx] <- anchor_idx

pair_names <- names(pair_multidesign)
node_counts <- vapply(pair_multidesign, function(dom) nrow(dom$x), integer(1))

hd_pair <- multidesign::hyperdesign(pair_multidesign)
```

## Running `parrot`

```{r run-parrot}
parrot_fit <- parrot(
  hd_pair,
  anchors = anchor_id,
  ncomp = 8,
  sigma = 0.2,
  lambda = 0.05,
  tau = 0.01,
  alpha = 0.3,
  gamma = 0.1,
  solver = "sinkhorn",
  max_iter = 80,
  tol = 1e-5,
  use_cpp = FALSE
)

str(parrot_fit, max.level = 1)
```

### Anchor and Class Diagnostics

The transport plan returned by `parrot()` is a soft correspondence between the two
domains.  We use it to derive a hard assignment (row-wise argmax) and inspect how
well the anchors and latent classes are preserved.  Labels are used only for
post-hoc diagnostics; the optimisation itself only sees the anchor IDs.

```{r parrot-diagnostics}
transport <- parrot_fit$transport_plan
assignment <- apply(transport, 1, which.max)

ref_design <- pair_multidesign[[1]]$design
cmp_design <- pair_multidesign[[2]]$design

class_accuracy <- mean(ref_design$condition == cmp_design$condition[assignment])
anchor_rows <- which(!is.na(ref_design$anchor_id))
anchor_accuracy <- mean(
  ref_design$anchor_id[anchor_rows] ==
    cmp_design$anchor_id[assignment[anchor_rows]],
  na.rm = TRUE
)

tibble(
  metric = c("Class agreement", "Anchor recovery"),
  value = c(class_accuracy, anchor_accuracy)
)
```

### Visualising the Transport Plan

```{r parrot-heatmap, fig.width = 6, fig.height = 5}
transport_norm <- as.matrix(transport / max(transport))
transport_df <- as_tibble(transport_norm, .name_repair = ~ paste0("V", seq_along(.))) %>%
  mutate(source = dplyr::row_number()) %>%
  pivot_longer(-source, names_to = "target", values_to = "weight") %>%
  mutate(target = as.integer(sub("V", "", target)))

subset_df <- dplyr::filter(transport_df, source <= 25, target <= 25)

 ggplot(subset_df, aes(x = target, y = source, fill = weight)) +
  geom_tile() +
  scale_fill_gradient(low = "#f0f0f0", high = "#0868ac") +
  labs(title = "PARROT transport plan (top-left 25 x 25 block)",
       x = paste(pair_names[2], "nodes"),
       y = paste(pair_names[1], "nodes"),
       fill = "weight") +
  theme_minimal()
```

### Barycentric Projection of Domain 2

Applying the transport plan to domain 2 (computing `S %*% X_2`) yields a
barycentric embedding in the coordinate system of domain 1.  This plot compares
those barycentric points to the original domain-1 positions, highlighting how
PARROT warps the second network while preserving class structure.

```{r parrot-barycentric, fig.width = 6, fig.height = 4}
X1 <- pair_multidesign[[1]]$x
X2 <- pair_multidesign[[2]]$x
X2_bary <- transport %*% X2

bary_df <- tibble(
  x = c(X1[, 1], X2_bary[, 1]),
  y = c(X1[, 2], X2_bary[, 2]),
  domain = rep(c("domain1", "barycentric"), each = nrow(X1)),
  condition = rep(ref_design$condition, 2)
)

 ggplot(bary_df, aes(x = x, y = y, colour = condition, shape = domain)) +
  geom_point(alpha = 0.85) +
  labs(title = "PARROT barycentric projection of domain 2",
       x = "Feature 1", y = "Feature 2") +
  theme_minimal()
```

## Summary

- `parrot()` blends feature distances with position-aware RWR descriptors and
  anchors to produce a soft alignment.
- On this benchmark pair a handful of anchors is enough to correctly recover the
  class structure and all provided anchors, and the barycentric projection shows
  domain 2 collapsing neatly onto domain 1.
- The same dataset underpins the `cone_align`, `gpca_align`, and `kema`
  vignettes, enabling like-for-like comparison across alignment strategies.

## Multi-domain consensus and diagnostics (align_many)

For three or more domains, you can use the generic orchestrator `align_many()`
with the PARROT adapter and a global, cycle-consistent permutation consensus.
Diagnostics such as edge and cycle residuals and entropy metrics are returned.

```r
# Example (not run): align 3 domains and inspect diagnostics
# (eval=FALSE to keep vignette light)

# pick three domains from the benchmark
triplet <- alignment_benchmark$domains[1:3]
X_list <- lapply(triplet, `[[`, "x")

# Run PARROT across all three with spectral permutation consensus
res <- align_many(
  domains = X_list,
  algo = parrot_aligner(),
  graph = "auto",
  sync = list(perm = list(mode = "spectral", rounding = "sinkhorn"))
)

# Inspect diagnostics
res$diagnostics$edge_residuals           # per-edge constraint residuals
res$diagnostics$mean_edge_residual       # global mean
res$diagnostics$cycle_residuals          # per-triangle residuals (if any)
res$diagnostics$entropy                  # per-domain soft-assignment entropy

# Barycentric embeddings to the consensus for domain 1
Z1 <- res$embeddings[[1]]

# Visualise cycle-consistency as a heatmap (min over paths)
library(ggplot2)
p_cc <- plot_cycle_consistency(res$diagnostics, aggregate = "min")
print(p_cc)

# Visualise edge residuals as a heatmap
p_er <- plot_edge_residuals_heatmap(res$diagnostics)
print(p_er)

# Summarize worst edges (top 5 by residual)
summarize_bad_edges(res$diagnostics, top = 5)
```
```
