---
title: 'Quickstart: Building a hyperdesign and running alignments'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart: Building a hyperdesign and running alignments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(params$family))
}
```

You have data from multiple sources — different instruments, subjects, or
conditions — and you want to project them into a shared space where samples
from the same class land near each other regardless of origin. That is what
**manifoldalign** does.

This vignette gets you from raw matrices to an aligned embedding in under
a minute.

## Setup

```{r setup}
library(manifoldalign)
library(multidesign)
library(ggplot2)
```

## Create a hyperdesign

A `hyperdesign` is the multi-domain container that every alignment function
in the package accepts. You build one from a list of data matrices plus
their associated design metadata.

We will use the built-in `alignment_benchmark` dataset, which provides three
domains with two latent classes and moderate domain-specific transformations.

```{r build-hyperdesign}
bench <- manifoldalign::alignment_benchmark

domain_list <- lapply(bench$domains, function(dom) {
  multidesign(dom$x, dom$design)
})

hd <- hyperdesign(domain_list)
hd
```

Each domain has `r nrow(domain_list[[1]]$x)` samples and its own feature
dimensionality. The `condition` column in each design frame holds the class
labels.

## Align with KEMA

`kema()` projects all domains into a shared latent space using kernel
manifold alignment. Pass the label column via `y` so the algorithm can
build within- and between-class graphs.

```{r run-kema}
kema_fit <- kema(
  hd,
  y = condition,
  ncomp = 2,
  knn = 3,
  sigma = 0.8,
  solver = "exact",
  preproc = multivarious::pass()
)
```

The result contains aligned scores for all domains stacked row-wise.

```{r plot-kema, echo = FALSE, fig.cap = "KEMA projects three domains into a shared 2D space. Points from the same class cluster together despite originating from different domains."}
domain_names <- names(domain_list)
domain_sizes <- vapply(domain_list, function(d) nrow(d$x), integer(1))

scores <- as.data.frame(as.matrix(kema_fit$s)[, 1:2])
colnames(scores) <- c("comp1", "comp2")
scores$domain <- rep(domain_names, times = domain_sizes)
scores$condition <- rep(bench$labels, length(domain_names))

ggplot(scores, aes(comp1, comp2, colour = condition, shape = domain)) +
  geom_point(size = 2.2, alpha = 0.85) +
  labs(x = "Component 1", y = "Component 2")
```

Classes separate cleanly, and the three domains overlap — alignment is
working.

## Align with GPCA

`gpca_align()` takes the same hyperdesign and produces a linear (non-kernel)
shared embedding. It is faster than KEMA on large datasets where kernel
matrices become expensive.

```{r run-gpca}
gpca_fit <- gpca_align(
  hd,
  y = condition,
  ncomp = 2,
  preproc = replicate(length(domain_list), multivarious::pass(), simplify = FALSE)
)
```

```{r plot-gpca, echo = FALSE, fig.cap = "GPCA alignment: a linear alternative that scales better to high-dimensional data."}
scores_gpca <- as.data.frame(as.matrix(gpca_fit$s)[, 1:2])
colnames(scores_gpca) <- c("comp1", "comp2")
scores_gpca$domain <- rep(domain_names, times = domain_sizes)
scores_gpca$condition <- rep(bench$labels, length(domain_names))

ggplot(scores_gpca, aes(comp1, comp2, colour = condition, shape = domain)) +
  geom_point(size = 2.2, alpha = 0.85) +
  labs(x = "Component 1", y = "Component 2")
```

## Building a hyperdesign from scratch

If you don't have a packaged dataset, build one from plain matrices:

```{r from-scratch}
set.seed(42)
X1 <- matrix(rnorm(80 * 5), 80, 5)
X2 <- matrix(rnorm(80 * 3), 80, 3)
labels <- rep(c("A", "B"), each = 40)

hd_custom <- as_hyperdesign(
  list(X1, X2),
  labels = list(labels, labels),
  label_name = "condition",
  domain_names = c("source", "target")
)

hd_custom
```

This `hyperdesign` works with every alignment function in the package.

## What's next?

Each alignment method has its own vignette with deeper examples:

- `vignette("kema-overview")` — full/regression/scalable KEMA solvers
- `vignette("gpca-align")` — graph sparsification and label balancing
- `vignette("cone-align")` — unsupervised graph alignment
- `vignette("parrot-align")` — optimal transport with anchors
- `vignette("fpgw_tutorial")` — fused partial Gromov-Wasserstein
- `vignette("performance-overview")` — method comparison table
